An Experimental Analysis of the Influence of the Precision of Parameters in an Artificial Neural Network

### 2025-07-16 paper/main.tex
fertig

\subsection*{Quantization During Training}
Several studies have explored training neural networks with reduced precision
directly. Notable contributions include:
\begin{itemize}
	\item \textbf{Mixed Precision Training}, which leverages float16 arithmetic for speed while maintaining float32 master weights~\cite{micikevicius2017mixed}.
	\item \textbf{DoReFa-Net}, which quantizes weights, activations, and gradients to arbitrary bitwidths~\cite{zhou2018dorefa}.
	\item \textbf{Q-BERT}, a quantized BERT model that maintains accuracy using only 4-bit weights~\cite{shen2020qbert}.
\end{itemize}

@@

\newpage
\section{Results}
In the first experiment, every floating-point tensor that is used inside the
training loop was converted to a lower precision data type. This therefore
includes the weights and biases, activations, gradients, optimizer states, etc.
The types tried were float64, float32, and float16.

\subsection*{Accuracy Across Bitwidths}
Figure~\ref{fig:modelAccTrainAndInf} shows how the accuracy changed after the
models were trained on the different precision settings. Keep in mind that the
y-axis is not starting from zero, so the absolute height differences are
amplified visually, though the delta is only around $.1$ percent. Re-training
the models and re-evaluating them has also lead to the accuracies being
``reversed'', meaning the float64 model had the least accuracy. It is safe to
assume that the model accuracy has neither decreased, nor increased.
\begin{figure}[H]
	\centering
	\includegraphics[width=0.85\textwidth]{figures/trainingAndInference.png}
	\caption{Model accuracy evaluated after training with the different precisions.}\label{fig:modelAccTrainAndInf}
\end{figure}

@@

\subsection*{Inference-Only Testing}
This third experiment was conducted to test a possible post-training precision
reduction. For this, the model was trained on the highest precision (float64)
and then quantized at the inference stage.
Figure~\ref{fig:inferenceOnlyTesting} shows that the accuracy of the baseline
(float64 training AND inference) is maintained as the precision is decreased at
the later stages. In contrast to the \textit{Accuracy Across Bitwidths}
experiment, there are no fluctuations at all because for the lower precisions,
there was no new model trained.
\begin{figure}[H]
	\centering
	\includegraphics[width=0.85\textwidth]{figures/inferenceOnly.png}
	\caption{Inference accuracy using float64-trained weights followed by quantization.}\label{fig:inferenceOnlyTesting}
\end{figure}

\subsection*{Automatic Mixed Precision and Gradient Clipping}
Automatic Mixed Precision (AMP) accelerates deep learning training by using
both float16 and float32 data types reducing memory usage and increasing
computational speed. However, float16’s limited dynamic range can cause
numerical instability, especially during backpropagation. To address this,
Gradient Clipping is used to cap the magnitude of gradients (either globally or
per element) preventing extreme updates that could hinder convergence.
Together, AMP and gradient clipping enable efficient, stable training by
balancing performance and numerical precision.

@@

\section{Summary}
% \begin{itemize}
% 	\item \textbf{16-bit is enough for this classification task}: End-to-end float16 preserved the accuracy.
% 	\item \textbf{Three decimals suffice}: Storing weights down to fractional length $3$ caused no visible accuracy loss.
% \end{itemize}

% \section{Discussion}
% The results of the first experiments have shown that either the selected task (MNIST
% classification) was not complex enough to be affected by the decrease in accuracy, or the
% accuracy would need to be decreased even further than float16. Otherwise, the results
% generally indicate that networks trained and evaluated with as low as 16-bit precision can
% achieve comparable performance to their high-precision counterparts. Moreover, rounding
% pretrained weights to just 3 decimal digits preserves full accuracy, opening the door to
% aggressive post-training quantization.This is particularly relevant for edge devices and
% real-time applications where memory and compute constraints are more relevant.

>>>>

\subsection*{Quantization During Training}
Several studies have explored training neural networks with reduced precision
directly. Notable contributions include:
\begin{itemize}
	\item \textbf{Mixed Precision Training}, which leverages FP16 arithmetic for speed while maintaining FP32 master weights~\cite{micikevicius2017mixed}.
	\item \textbf{DoReFa-Net}, which quantizes weights, activations, and gradients to arbitrary bitwidths~\cite{zhou2018dorefa}.
	\item \textbf{Q-BERT}, a quantized BERT model that maintains accuracy using only 4-bit weights~\cite{shen2020qbert}.
\end{itemize}

@@

\newpage
\section{Results}
In the first experiment, every floating-point tensor that is used inside the
training loop was converted to a lower precision data type. This therefore
includes the weights and biases, activations, gradients, optimizer states, etc.
The types tried were FP64, FP32, and FP16.

\subsection*{Accuracy Across Bitwidths}
Figure~\ref{fig:modelAccTrainAndInf} shows how the accuracy changed after the
models were trained on the different precision settings. Keep in mind that the
y-axis is not starting from zero, so the absolute height differences are
amplified visually, though the delta is only around $.1\%$. Re-training the
models and re-evaluating them has also lead to the accuracies being
``reversed'', meaning the FP64 model had the least accuracy. It is safe to
assume that the model accuracy has neither decreased, nor increased.
\begin{figure}[H]
	\centering
	\includegraphics[width=0.85\textwidth]{figures/trainingAndInference.png}
	\caption{Model accuracy evaluated after training with the different precisions.}\label{fig:modelAccTrainAndInf}
\end{figure}

@@

\subsection*{Inference-Only Testing}
This third experiment was conducted to test a possible post-training precision
reduction. For this, the model was trained on the highest precision (FP64) and
then quantized at the inference stage. Figure~\ref{fig:inferenceOnlyTesting}
shows that the accuracy of the baseline (FP64 training AND inference) is
maintained as the precision is decreased at the later stages. In contrast to
the \textit{Accuracy Across Bitwidths} experiment, there are no fluctuations at
all because for the lower precisions, there was no new model trained.
\begin{figure}[H]
	\centering
	\includegraphics[width=0.85\textwidth]{figures/inferenceOnly.png}
	\caption{Inference accuracy using FP64-trained weights followed by quantization.}\label{fig:inferenceOnlyTesting}
\end{figure}

\subsection*{Automatic Mixed Precision and Gradient Clipping}
Automatic Mixed Precision (AMP) accelerates deep learning training by using
both FP16 and FP32 data types reducing memory usage and increasing
computational speed. However, FP16’s limited dynamic range can cause numerical
instability, especially during backpropagation. To address this, Gradient
Clipping is used to cap the magnitude of gradients (either globally or per
element) preventing extreme updates that could hinder convergence. Together,
AMP and gradient clipping enable efficient, stable training by balancing
performance and numerical precision.

@@

\section{Summary}
% \begin{itemize}
% 	\item \textbf{16-bit is enough for this classification task}: End-to-end FP16 preserved the accuracy.
% 	\item \textbf{Three decimals suffice}: Storing weights down to fractional length $3$ caused no visible accuracy loss.
% \end{itemize}

% \section{Discussion}
% The results of the first experiments have shown that either the selected task (MNIST
% classification) was not complex enough to be affected by the decrease in accuracy, or the
% accuracy would need to be decreased even further than FP16. Otherwise, the results
% generally indicate that networks trained and evaluated with as low as 16-bit precision can
% achieve comparable performance to their high-precision counterparts. Moreover, rounding
% pretrained weights to just 3 decimal digits preserves full accuracy, opening the door to
% aggressive post-training quantization.This is particularly relevant for edge devices and
% real-time applications where memory and compute constraints are more relevant.
